# Complex Datatypes and File Reading

```{r setup-ch-03, include=FALSE}
library(dplyr)
library(ggplot2)
```


## Factors: Nominal & Ordinal scales

Although factors are not actually a complex datatype but very much one of the five base types in R, I saved them because they have some more complex and sometimes puzzling behaviour. 

:::definitions
Factors represent different discrete levels of a variable - the _nominal_ and _ordinal_ scales known from statistics. 
:::

For instance:  

- eye color (brown, blue, green)
- weight class (underweight, normal, obese)
- autism spectrum (none, minimal, heavy) 


#### Factor creation {-}

Factors are used to represent data in nominal and ordinal scales. **_Nominal_** has no order (e.g. eye color). **_Ordinal_** has order (e.g. autism spectrum), but can not be calculated with, other than ordering from high to low. No _distance_ is defined between separate levels. The following functions are used to create factors:

- `factor()`: constructor function, from factor, character, numeric or logical 
- `as.factor()`: coercion function, from factor, character, numeric or logical 
- `cut()`: conversion function from numeric vector

So what is the difference between `factor()` and `as.factor()`? Function `as.factor()` is a _wrapper_ for `factor()`. The difference lies in behaviour when the input is a factor itself: `factor` will omit unused levels. Besides this, `as.factor()` does not specify the arguments for labels and levels.

```{r, results='hold'}
x <- factor(c("a", "b"), levels = c("a", "b", "c"))
x
factor(x)
as.factor(x)
```

Suppose you have surveyed the eye color of your class room and found these values

```{r eye-color-1, tidy=FALSE}
eye_colors <- c("green", "blue", "brown", "brown", "blue",
    "brown", "brown", "brown", "blue", "brown", "green",
    "brown", "brown", "blue", "blue", "brown")
```

Next you would like to plot or tabulate these findings. Simply plotting gives an error:

```{r eye-color-2, error=TRUE, out.width='80%', fig.asp=.75, fig.align='center', results='hold'}
plot(eye_colors)
```

However, plotting a character vector converted to a factor is easy

```{r eye-color-3, out.width='80%', fig.asp=.75, fig.align='center'}
eye_colors <- as.factor(eye_colors)
plot(eye_colors)
```

Factors are also really easy to tabulate and filter

```{r eye_-olor-4}
table(eye_colors)
sum(eye_colors == "blue")
```

#### Levels, Labels and Ordering {-}

When working with ordinal scales, defining the order of the factors (levels) is crucial. By default, R uses the _natural ordering_ which means it will stick to either numerical (`numeric`, `integer` and `logical`) or alphabetical ordering (`character`). When you want a different ordering you need to specify this. You can even define missing levels, as shown in the following example.

```{r class-sizes-1, out.width='80%', fig.asp=.75, fig.align='center'}
classSizes <- factor(c("big", "small", "huge", "huge", 
    "small","big","small","big"),
    levels = c("small", "normal", "big", "huge"),
    ordered = TRUE) #make it an ordinal scale!
plot(classSizes)
```

When you have factor, you can do a -limited- set of calulations with it. However, comparators only work with ordinal scale. As with all equality tests, `sum()` works as well:

```{r class-sizes-2}
classSizes < "big" ## only with in Ordinal scale
sum(classSizes == "huge") 
```

#### Convert existing factors {-}

When you already have an unordered factor, you can make it ordered by using the function `ordered()` together with a fvector specifying the levels.

```{r class-sizes-3}
classSizes <- factor(c("big", "small", "huge", "huge",
    "small", "big", "small", "big"))
classSizes <- ordered(classSizes,
                    levels = c("small", "big", "huge"))
classSizes
```

#### When calculations get corrupted {-}

Especially when a factor consists of numeric levels, calculations can get your mind screwed big time: 

```{r calc-on-factor-error}
x <- factor(c(3, 4, 5, 4))
x + 1
as.integer(x) + 1
as.integer(levels(x)) + 1
```

The only way to get the numbers back with numeric factors is by using this trick

```{r}
x
as.integer(levels(x))[x]
```

But this makes for really unintelligible code so try to prevent this at all costs!

#### The power of factors {-}

Factors are used all the time e.g. for defining treated/untreated. That's why R knows how to deal with them so well:

```{r factor-plotting, out.width='80%', fig.asp=.75, fig.align='center'}
with(ChickWeight, plot(weight ~ Diet))
```

You will see many many examples of factors in the subsequent chapters of this ebook.

But it's time to switch to a better plotting platform. Here is the same plot with the `ggplot2` library:

```{r factor-plotting-ggplot, out.width='80%', fig.asp=.75, fig.align='center'}
ggplot(data = ChickWeight,
       mapping = aes(x = Diet, y = weight, fill = Diet)) +
  geom_boxplot() +
  theme(legend.position = "none")
```

From here on, I will only use the base R plotting system for quick-and-dirty visualizations, just like you should.

## Matrices are vectors with dimensions

We will not detail on them in this course, only this one small paragraph. This does not mean they are not important, but they are just not the focus here. Some functions require or return a matrix so you should be aware of them.


```{r matrix_demo}
m <- matrix(1:10, nrow = 2, ncol = 5) 
m
v <- 1:10
dim(v) <- c(2, 5)
v
```

You can convert to and from dataframes using `as.data.frame()` and `as.matrix()`. From and to simple vectors is removing or adding dimension attributes: `dim(m) <- NULL`.  


## Lists

:::definitions
A list is an **_ordered collection of objects_**. These objects can be of any type: vectors (also of different types), dataframes, lists.
:::

#### List creation {-}

Create a list with or without element names:  

- `list(element1, element2, ...)`  
- `list(name1 = element1, name2 = element2, ...)`

Without names:

```{r list-action1, results='hold'}
x <- c(2, 3, 1);  y <- c("foo", "bar")
l <- list(x, y)
l
```

With names:

```{r list-named-elements-1}
x <- c(2, 3, 1)
y <- c("foo", "bar")
l <- list("numbers" = x, "words" = y)
l
```

This is the preferred way to create and use them because it gives you more and easier ways to access its elements and it makes for much more reradable code. That's why you will only see lists with named elements from here on.

#### Making selections on lists {-}

Accessing named elements can be done in three ways:

- By index, within double or single brackets: `[[<index>]]` or `[<index>]`
- By name of the element, within double or single brackets: `[[<name>]]` or `[<name>]`
- By name of the element, using the dollar sign on the list name: `$<name>`

Here are all three:

```{r list-named-elements-2, results='hold'}
l[[2]]        # index
l[["words"]]  # name of element with double brackets
l$words       # name of element with dollar sign
```

:::attention
Single bracket selection on a list returns a list; double brackets and `$` return a vector.
:::

```{r list-brackets-1, results='hold'}
l[2]
l[[2]]
l$words
```

In R, selections are often **_chained_**. In the following example the second vector element of the second list element is selected.

```{r list-action2, results = 'hold'}
l
l[[2]][2] 
```

When you need multiple elements of a list, use **_single brackets_**. Remember: single brackets return a list; that's why you need single brackets here.

```{r}
l[c(1,2,1)]
```

Accessing named elements has its limitations. You can not use a variable in combination with the dollar sign selector.

```{r list-named-elements-3}
select <- "words"
l[[select]] ## OK
l$select ##fails - no element with name "select"
```

Chaining of selectors can become awkward, as this example demonstrates.

```{r list-brackets-2}
l[2]["words"][1]$words  ## mind****
```


## Dataframes

:::definitions
A dataframe is an **_ordered collection of vectors_**. These vectors can have _differing types_ but must have _equal lengths_.
:::

A dataframe is very similar to the square grid-like structures you have probably worked with in Excel. Variables are in columns in which all elements are of the same type. Examples (observations) are in rows - they _can_ have differing types.

Dataframes can be constructed using the `data.frame()` function in the same way as the `list` function:  
`data.frame(column1 = vector1, column2 = vector2, ...)` 

Here is a first example.

```{r data-frame-1, tidy=FALSE}
geneNames <- c("P53","BRCA1","VAMP1", "FHIT")
sig <- c(TRUE, TRUE, FALSE, FALSE)
meanExp <- c(4.5, 7.3, 5.4, 2.4)
genes <- data.frame(
    "name" = geneNames,  
    "significant" = sig,  
    "expression" = meanExp)  
genes
```

By wrapping this dataframe in a more modern variant of `data.frame`, the `tibble`, you can see the structure: each column has a single datatype but rows can have differing types for neighboring fields.

```{r}
dplyr::as_tibble(genes)
```


### Selections on dataframes

Making selections on dataframes is not very surprising when you already know how to do it with vectors and lists. There is only one extension. The fact that it is a square grid-like structure makes it possible to add an extra way of making selections: combining rows and column selections as subgrids. This section extensively reviews all means of making selections.

This is a summary:  

- Select a single column using `$` will return a vector
- Selecting with double brackets `[[<name>]]` or `[[<index>]]` will return a vector
- Selecting with single brackets `[<name>]` or `[<index>]` will return a dataframe
- Selecting with row-and-column coordinates `[row_selection, col_selection]` returns either a vector or a dataframe, depending on the selection made. Here, `row_selection` and `col_selection` can be 
    - a numerical vector of length 1 or more
    - a logical vector of length 1 or more
    - empty (to select all rows/columns)
    
Here follow a few examples.

```{r data-frame2, prompt = TRUE}
genes[2,1] 		    #row 2, column 1
genes[2, 1:2]     #row 2, columns 1 and 2
genes[2, c(1, 3)] #row 2, column 1 and 3
genes$name 	      #column "name"
genes[, c("name", "expression")]	#columns "name" and "expression", all rows
genes[, 1:2] 	    #columns 1 and 2, all rows
genes[1:2, ]      #row 1 and 2, all columns
```

As with vectors and lists, R will cycle selectors, and you can select an element as often as you want.

```{r data_frame4}
genes[c(T, F), 1]	#every uneven row, column 1
genes[c(1, 1, 1, 2), ]	#three times row 1 and row 2
```

A dataframe is much like a list, but not entirely equal:

```{r data-frame5}
genes[["name"]] ## select column w. double brackets
class(genes) ## it is NOT a list though
str(genes)
```

:::attention
A dataframe can usually be treated as if it was a list, but a list can rarely be treated as a dataframe.
:::


#### Selections with `subset()` {-}

:::info
This section is optional; we are going to proceed with the `dplyr` package that makes these operations easier.
:::

Function `subset()` can serve as alternative to "bracket-based" selections (`[ , ]`).
You can use `subset()` to make both **column** and **row selections**. Here, using `subset = `, the rows are selected for which Solar.R is available using the `is.na()` function.

```{r}
head(subset(airquality, subset = !is.na(Solar.R)))
```

Note that you don't even need to use quotes for column names.

Select columns only with the `select = ` argument.

```{r}
head(subset(airquality, select = c(Ozone, Solar.R)))
```

Of course, you can combine row and colum selection:

```{r}
head(subset(airquality, 
            subset = !is.na(Solar.R), 
            select = c(Ozone, Solar.R)))
# shorthand notation
#subset(airquality, Day == 1, select = -Temp)
```

`subset()` can be used more sophisticated; however we are going to see `subset()` on steroids in the next course: the functions in package `dplyr`.


## Importing dataframes using `read.table()`

Usually your data comes from file, loaded into memory as a `data.frame`. The most common data transfer- and storage format is text. The text file will have column separators that can be any of a wide range of characters called **_delimiters_**, but tab- or comma-delimited fields are most common.  

Here is an example dataset in a file (see [whale_selenium.txt](data/whale_selenium.txt)) where the separator is a space character:

    whale liver.Se tooth.Se  
    1 6.23 140.16  
    2 6.79 133.32  
    3 7.92 135.34  
    ...  
    19 41.23 206.30  
    20 45.47 141.31  

To load this data into an R session you can use the function `read.table()`. Let's try

```{r data-frame-io1}
whale_selenium <- read.table("data/whale_selenium.txt")
head(whale_selenium) # first rows
str(whale_selenium) # structure
```

That is not entirely correct: all columns are imported as a factor while obviously they should be numeric. The cause of this is that, when loading the data,  

- there is no special consideration for the header line
- the separator is assumed to be a space
- the decimal is assumed to be a dot "."

(and some more assumptions)

Therefore, to read a file correctly, you have to specify its format in every detail. in this case,  

- the first line is a header with column names
- the first column contains the row names

Here is a new attempt with some **_format specifications_**:


```{r data-frame-io2}
whale_selenium <- read.table(
    file = "data/whale_selenium.txt",
    header = TRUE,
    row.names = 1)
```

Before proceeding with your data, you should always perform some checks. Several helper methods exist for this purpose:

- `head()` shows you the first `n` lines
- `str()` gives you a structure description: what types have the columns and what dimension does the data frame have?
- `summary()` gives you a 6-number sumary of the data

```{r data-frame-inspect, prompt = TRUE}
head(whale_selenium, n=4) 
str(whale_selenium) 
summary(whale_selenium) 
```

### File I/O in detail

Whatever the contents of a file, you always need to address (some of) these questions:

- Are there comment lines at the top?
- Is there a header line with column names?
- What is the column separator?
- Are there quotes around character data?
- How are missing values encoded?
- How are numeric values encoded?
- Are there dates (a special challenge)
- What is the type in each column?
    - character / numeric / factor / date-time


#### Some `read.table()` arguments {-}

| arg              | specifies          | example           | 
|------------------|--------------------|-------------------| 
| **sep**          | field separator    | sep = ":"         | 
| **header**       | is there a header  | header = F        | 
| **dec**          | decimal format     | dec = ","         | 
| **comment.char** | comment line start | comment.char = "" | 
| **na.strings**   | NA value           | na.strings = "-"  | 
| **as.is**        | load as character  | as.is = c(1,4)    | 
| **stringsAsFactors** | load strings as factors | stringsAsFactors = F |


#### The data reading workflow {-}

Always apply this sequence of steps and repeat until you are satisfied with the result:

1. `read.table()` with arguments that seem OK
2. Check the result at least with `str()` and `head()` and verify that the columns have the correct data type.
    - Factors where numeric expected indicate missed "NA" values!
3. Adjust the read.table parameters
4. Rinse and repeat


## Simple dataframe manipulations

There are various other helper methods the you can use to inspect the contents and nature of your dataframe columns and rows:

- `dim()` gives the rows and columns
- `ncol()` gives the number of columns
- `nrow()` gives the number of rows
- `names()` gives the column names (synonym to `colnames()`)
- `rownames()` gives the row names

But visualizing your data speaks more than a thousand words of course.


```{r data-frame-io3, out.width='80%', fig.asp=.75, fig.align='center'}
ggplot(data = whale_selenium,
       mapping = aes(x = liver.Se, y = tooth.Se)) +
  geom_point() +
  geom_smooth(method = "lm", formula = "y~x", se = F) +
  xlab("Liver Selenium") +
  ylab("Tooth Selenium")
  
```


### Changing column names {-}

Sometimes the existing column names are not convenient to work with (unclear, too long etc.). In that case it may be a good idea to change the column names. To do this you can use either `names()` or `colnames()`.

```{r rename-columns}
names(whale_selenium) <- c("liver", "tooth")
head(whale_selenium, n=2)
##or
colnames(whale_selenium) <- c("pancreas", "colon")
head(whale_selenium, n=2)
```

```{r echo = FALSE, results = "hide"}
colnames(whale_selenium) <- c("liver", "tooth")
```


### Adding columns {-}

You can add a single column by simply specifying its name and the value(s) to be attached.

```{r add_columns}
## add simulated stomach data
whale_selenium$stomach <- rnorm(nrow(whale_selenium), 42, 6) 
head(whale_selenium, n=2)
```

Alternatively, use `cbind`. It is a bit more versatile because you can add multiple columns at once.

```{r add-columns2}
cbind(whale_selenium, "brain" = c(1, 0)) #cycled values!
```


### Adding rows: `rbind()` {-}

Adding rows to a dataframe is similar. There is however a constraint: the column names of both dataframes need to match for this operation to succeed.

```{r add_rows1}
my_data1 <- data.frame(colA = 1:3, colB = c("a", "b", "c"))
my_data2 <- data.frame(colA = 4:5, colB = c("d", "e"))
my_data_complete <- rbind(my_data1, my_data2)
my_data_complete
```


