# Built-in Functions


```{r, include=FALSE}
options(digits = 3)
library(ggplot2)
knitr::opts_knit$set(cache = TRUE,
                     tidy = TRUE,
                     tidy.opts = list(blank = FALSE, width.cutoff = 60))
```


This chapter deals with the most-used functions that R has to offer. A clean installation of R already contains hundreds of functions; see [here](https://stat.ethz.ch/R-manual/R-devel/library/base/html/00Index.html) for a complete listing of functions in the `base` package.
Obviously, this eBook is not the platform to discuss these exhaustively. Therefore, I chose to only deal with the ones I use regularly.  
Besides that, the tidyverse packages `tidyr` and `dplyr` offer functionality that are better alternatives for many base R functions. These packages are discussed in later chapters. 


## Descriptive statistics

R provides a wealth of descriptive statistics functions. The most important ones of them are listed below. The ones with an asterisk are described in more detail in following paragraphs.

| function        |  purpose             | 
|-----------------|----------------------| 
| `mean( )`       | mean                 | 
| `median( )`     | median               | 
| `min( )`        | minimum              | 
| `max( )`        | maximum              | 
| `range( )`      | min and max          | 
| `var( )`        | variance s^2         | 
| `sd( )`         | standard deviation s | 
| `summary( )`    | 6-number summary     | 
| `quantile( )` * | quantiles            | 
| `IQR( )`  *     | interquantile range  | 


#### The `quantile()` function {-}

This function gives the data values corresponding to the specified quantiles. The function defaults to the quantiles `0%  25%  50%  75% 100%`: these are the _quartiles_ of course.

```{r quantiles-1}
quantile(ChickWeight$weight)
quantile(ChickWeight$weight, probs = seq(0, 1, 0.2))
```

#### Interquantile range `IQR()` {-}

Function `IQR()` gives the range between the 25% and 75% quantiles.

```{r quantiles-2}
IQR(ChickWeight$weight)
## same as
quantile(ChickWeight$weight)[4] - quantile(ChickWeight$weight)[2]
## same as
diff(quantile(ChickWeight$weight, probs = c(0.25, 0.75)))
```

#### `boxplot()` is `summary()` visualized {-}

Boxplot is a graph of the 5-number summary, but `summary()` also gives the mean 

```{r box_demo, fig.height=2.5, fig.width=3.5}
summary(ChickWeight$weight)
boxplot(ChickWeight$weight)
```


## General purpose functions

### Dealing with NAs

Dealing with NA is a **_very big thing_**. When you work with external data there is always the possibility that some values will be missing.

You should be aware that it is not possible to test for `NA` values (Not Available) values in any other way; using `==` will simply return `NA`:

```{r dealing-with-na-1}
x <- NA
x == NA
```

Important functions for dealing with that are `na.omit()` (`drop_na()` in dplyr) and `complete.cases()`. The first drops rows from a dataframe that contain at least one NA; the other returns a logical indicating which rows are complete; i.e. without any NA.  
Besides that, many functions in R have a (variant of) the `na.rm = ` argument. For instance, when the `sum()` function encounters an NA in its input vector, it will always return `NA`:

```{r dealing-with-na-2}
x <- c(1, 2, 3, NA)
sum(x)
sum(x, na.rm = TRUE)
```


### Convert numeric vector to factor: `cut()`

Sometimes it is useful to work with a factor instead of a numeric vector. For instance, when working with a Body Mass Index (bmi) variable it may be nice to split this into a factor for some analyses.
The function `cut()` is used for this.
Suppose you have the following fictitious dataset

```{r cut-demo-1, tidy = F}
## body mass index
bmi <- c(22, 32, 21, 37, 28, 34, 26, 29,
         41, 18, 22, 27, 32, 31, 26)
## year income * 1000 euros
income <- c(23, 14, 20, 13, 47, 15, 38, 29, 
            12, 25, 33, 24, 19, 42, 38)
my_data <- data.frame(bmi = bmi, income = income)
```

You can of course look at income as a function of bmi using a scatter plot:

```{r cut-demo-2, out.width='70%', fig.asp=.75, fig.align='center'}
with(my_data, plot(income ~ bmi))
```

But wouldn't it be nice to look at the bmi categories as defined by the WHO? To be able to do this, you need to split the numeric `bmi` variable into a factor using `cut()`.


```{r cut-demo-3, out.width='70%', fig.asp=.75, fig.align='center'}
my_data$bmi_class <- cut(bmi,
    breaks = c(0, 18.5, 25.0, 30.0, Inf), 
    right = F,
    labels = c("underweight", "normal", "overweight", "obese"),
    ordered_result = T)
with(my_data, boxplot(income ~ bmi_class))
```

The `breaks =` argument specifies the split positions; the `right = F` arguments specifies that the interval is _inclusive_ on the lower (left) boundary:

```{r cut-demo-4}
x <- c(2, 5, 10)
cut(x, breaks = c(0, 2, 5, 10), right = F)
cut(x, breaks = c(0, 2, 5, 10), right = T)
```

An interval written as `(5,10]` means it is from -but excluding- 5 to -but including- 10. 
Note that in the first example the last value (10) becomes NA because 10 is exclusive in that interval specification.


### Memory management {-}

When working with large datasets it may be useful to free some memory once in a while (i.e. intermediate results). Use `ls()` to see what is in memory; use `rm()` to delete single or several items: `rm(genes)`, `rm(x, y, z)` and clear all by typing `rm(list = ls())`


### File system operations {-}

Several functions exist for working with the file system:  

- `getwd()` returns the current working directory.
- `setwd(</path/to/folder>)` sets the current working directory.
- `dir()`, `dir(path)` lists the contents of the current directory, or of `path`.
- A `path` can be defined as `"E:\\emile\\datasets"` (Windows) or, on Linux/Mac using relative paths `"~/datasets"` or absolute paths `"/home/emile/datasets"`. 


### Glueing character elements: `paste()` {-}

Use `paste()` to combine elements into a string

```{r paste-demo-2}
paste(1, 2, 3)
paste(1, 2, 3, sep="-")
paste(1:12, month.abb)
```

There is a variant, `paste0()` which uses no separator by default. 

### A local namespace: `with()` {-}

When you have a piece of related code operating on a single dataset, use `with()` so you don't have to type its name all the time.

```{r with-demo, eval=FALSE}
with(airquality, {
  mdl <- lm(Solar.R ~ Temp)
  plot(Solar.R ~ Temp)
  abline(mdl)
})
```

Local variables such as `mdl` will not end up in the global environment. 


### Writing data to file {-}

To write a data frame, matrix or vector to file, use `write.table(myData, file="file.csv")`. Standard is a comma-separated file with both column- and row names, unless otherwise specified:  

- `col.names = F`
- `row.names = F`
- `sep = ";"`
- `sep = "\t" # tab-separated`

### Saving R objects to file {-}

Use the `save()` function to write R objects to file for later use.
This is especially handy with intermediate results in long-running analysis workflows.

```{r save-obejcts-to-file, eval = FALSE}
x <- stats::runif(20)
y <- list(a = 1, b = TRUE, c = "oops")
save(x, y, file = "xy.RData")
```


### Writing a plot to file {-}

Use one of the functions `png()`, `jpeg()`, `tiff()`, or `bmp()` for these specific file types. They have widely differing properties, especially with respect to file size.  
Use `width` and `height` to specify size. Default unit is pixels. Use other unit: `units = "mm"`

```{r writing-plot-to-file, eval = FALSE}
png("/path/to/your/file.png",
    width = 700, height = 350, units = "mm")
plot(cars)
dev.off() # don't forget this one!
```

A ggplot figure can be written to file like this:

```{r writing-ggplot-to-file, eval = FALSE}
png("myplot.png")
myplot <- ggplot(mtcars, aes(wt, mpg)) + 
  geom_point()
print(myplot)
dev.off()
```


