# Old school data mangling {#mangling}

```{r, include=FALSE}
options(digits = 3)
knitr::opts_knit$set(cache = TRUE,
                     tidy = TRUE,
                     tidy.opts = list(blank = FALSE, width.cutoff = 60))
library(ggplot2)
```

This chapter  deals with old-school data mangling using base R functions. These are presented for completeness' sake, and also because sometimes these functions are simply the most convenient compared to what dplyr has to offer.

Dataframes are ubiquitous in R-based data analyses. Many R functions and packages are tailored specifically for DF manipulations - you have already seen `cbind()`, `rbind()` and `subset()`.  
In this presentation, we'll explore a few new functions and techniques for working with DFs:

- `apply()`
- `lapply()`
- `sapply()`
- `tapply()`
- `aggregate()`
- `split()`


## The `apply()` family of functions

Looping with `for` may be tempting, but highly discouraged in R because its inefficient. Usually one of these functions will do it better (or of course the dplyr functions discussed below):

- `apply`: Apply a function over the "margins" of a dataframe - rows or columns or both
- `lapply`: Loop over a list and evaluate a function on each element; returns a list of the same length
- `sapply`: Same as lapply but try to simplify the result
- `tapply`: Apply a function over subsets of a vector (read: split with a factor)

There are more but these are the important ones.

#### `apply()`: Apply Functions Over Array Margins

- Suppose you want to know the means of all columns of a dataframe.
- `apply()` needs to know 
    1. what DF to apply to (`X`)
    2. over which margin(s) - columns and/or rows (`MARGIN`)
    3. what function to apply (`FUN`)

```{r apply_demo}
apply(X = cars, MARGIN = 2, FUN = mean) # apply over columns
```

Here, a function is applied to both columns and rows

```{r}
df <- data.frame(x = 1:5, y = 6:10)
minus_one_squared <- function(x) (x-1)^2
apply(X = df, MARGIN = c(1,2), FUN = minus_one_squared)
```

(Ok, that was a bit lame: `minus_one_squared(df)` does the same)

The Body Mass Index, or BMI, is calculated as $(weight / height ^ 2) * 703$ where weight is in pounds and height in inches. Here it is calculated for the build in dataset `women`.

```{r apply_demo2}
head(women, n=3)
women$bmi <- apply(X = women, 
                   MARGIN = 1, 
                   FUN = function(x) (x[2] / x[1]^2) * 703)
head(women, n=4)
```

#### Pass arguments to the applied function {-}

Sometimes the applied function needs to have other arguments passed besides the row or column. The `...` argument to `apply()` makes this possible (type `?apply` to see more info)

```{r}
# function sums and powers up
spwr <- function(x, p = 2) {sum(x)^p}
# a simple dataframe
df <- data.frame(a = 1:5, b = 6:10)
df
# spwr will use the default value for p (p = 2)
apply(X = df, MARGIN = 1, FUN = spwr) 
# pass power p = 3 to function spwr (argument names omitted)
apply(df, 1, spwr, p = 3) 
```

Note: The `...` argument works for all `..apply..` functions.

#### `lapply()`: Apply a Function over a List or Vector

Function `lapply()` applies a function to all elements of a list and returns a list with the same length, each element the result of applying the function

```{r lapply_demo1}
myNumbers = list(
    one = c(1, 3, 4), 
    two = c(3, 2, 6, 1), 
    three = c(5, 7, 6, 8, 9))
lapply(X = myNumbers, FUN = mean)
```


Here is the same list, but now with `sqrt()` applied. Notice how the nature of the applied function influences the result.

```{r lapply_demo2}
lapply(X = myNumbers, FUN = sqrt)
```


#### `sapply()`: Apply a Function over a List or Vector and Simplify

When using the same example as above, but with `sapply`, you get a vector returned. Note that the resulting vector is a named vector, a convenient feature of `sapply` 

```{r sapply_demo}
myNumbers = list(
    one = c(1, 3, 4),
    two = c(3, 2, 6, 1),
    three = c(5, 7, 6, 8, 9))
sapply(X = myNumbers, FUN = mean)
```

When the result can not be simplified, you get the same list as with `lapply()`:

```{r}
sapply(X = myNumbers, FUN = sqrt)
```

#### wasn't a dataframe also a list?

Yes! It is also list(ish). Both `lapply()` and `sapply()` work just fine on dataframes:

```{r df-is-also-list}
lapply(X = cars, FUN = mean)
sapply(X = cars, FUN = mean) 
```

By the way, sapply and lapply also work with vectors.

#### `tapply()`: Apply a Function Over a Ragged Array

What `tapply()` does is apply a function over subsets of a vector; it splits a vector into groups according to the levels in a second vector and applies the given function to each group.

```{r tapply_demo}
tapply(X = chickwts$weight, INDEX = chickwts$feed, FUN = sd)
```

## Other data mangling functions

#### `split()`: Divide into Groups and Reassemble

This is similar to `tapply()` in the sense that is uses a factor to split its first argument. But where `tapply()` splits a vector, `split()` splits a dataframe - into _list of dataframes_.
You use `split()` when a dataframe needs to be divided depending on the value of some grouping variable.  
Here we have the response of Treated (T) and Untreated (UT) subjects

```{r split_demo-1}
myData <- data.frame(
    response = c(5, 8, 4, 5, 9, 3, 6, 7, 3, 6, 5, 2),
    treatment = factor(
        c("UT", "T", "UT", "UT", "T", "UT", "T", "T", "UT", "T", "T", "UT")))
splData <- split(x = myData, f = myData$treatment)
str(splData)
boxplot(splData$T$response, splData$UT$response, 
        names = c("Treated", "Untreated"))
```

Note that this trivial example could also have been done with `boxplot(myData$response ~ myData$treatment)`.

Here you can see that `split()` also works with vectors.

```{r split-demo-2}
split(x = rnorm(10), f = rep(c("sick", "healthy"), each=5))
```

#### `aggregate()`: Compute Summary Statistics of Data Subsets

Splits the data into subsets, computes summary statistics for each, and returns the result in a convenient form.

```{r aggregate-demo-1}
aggregate(Temp ~ Month, 
          data = airquality, 
          FUN = mean)
```

Aggregate has two usage techniques:  

- with a formula:  
  **`aggregate(formula, data, FUN, ...)`**  

- with a list:  
  **`aggregate(x, by, FUN, ...)`**  

I really like `aggregate()`, especially the first form. That is, until I got to know the `dplyr` package.

Both forms of `aggregate()` will be demonstrated

#### Aggregate with formula {-}

The left part of the formula accepts one, several or all columns as dependent variables.

```{r aggregate-demo-2}
##two dependents
aggregate(cbind(Temp, Ozone) ~ Month, 
          data = airquality, 
          FUN = mean)
##all
aggregate(. ~ Month, 
          data = airquality, 
          FUN = mean)
```


The right part can also accept multiple independent variables


```{r aggregate-demo-3}
airquality$Temp_factor <- cut(airquality$Temp, 
                              breaks = 2, 
                              labels = c("low", "high"))
aggregate(Ozone ~ Month + Temp_factor, 
          data = airquality, 
          FUN = mean)
```

#### The `by = list(...)` form {-}

This is the other form of aggregate. It is more elaborate in my opinion because you need te spell out all vectors you want to work on.

```{r aggregate-demo-4}
aggregate(x = chickwts$weight, 
          by = list(feed = chickwts$feed), 
          FUN = mean)
```

Here is another example:

```{r aggregate-demo-5}
aggregate(x = airquality$Wind, 
          by = list(month = airquality$Month, temperature = airquality$Temp_factor), 
          FUN = mean)
```

But it is better to wrap it in `with()`:

```{r aggregate-demo-6, eval=FALSE}
with(airquality, aggregate(x = Wind, 
                           by = list(month = Month, temperature = Temp_factor), 
                           FUN = mean))
```

### Many roads lead to Rome

The next series of examples are all essentially the same. The message is: there is more than one way to do it!

```{r roads-to-rome-1}
aggregate(weight ~ feed, 
          data = chickwts, 
          FUN = mean)
```

same as

```{r roads-to-rome-2}
head(aggregate(x = chickwts$weight, 
               by = list(feed = chickwts$feed), 
               FUN = mean), n=3)
```

same as

```{r roads-to-rome-3, message=FALSE}
tapply(chickwts$weight, chickwts$feed, mean)
with(chickwts, tapply(weight, feed, mean))
```

same as

```{r roads-to-rome-4}
sapply(split(chickwts, chickwts$feed), function(x){mean(x$weight)})
```

And this is the topic of the next part of this chapter:

```{r roads-to-rome-5, message=FALSE}
library(dplyr)
group_by(chickwts, feed) %>% 
  summarise(mean_weigth = mean(weight))
```

