# Working with date-time data {#date-time}

```{r setup-stringr, include=FALSE}
options(digits = 3)
knitr::opts_knit$set(cache = TRUE,
                     tidy = TRUE,
                     tidy.opts = list(blank = FALSE, width.cutoff = 60))
library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)
```

## The `lubridate` package

### Parsing character data

When working in a lab you will often encounter time-measured data. For instance, growth curves, animal behavior observations etc.

The tidyverse package `lubridate` can help you with that. It is quite an expensive package; only the bare essentials will be dealt with here. For a bit more complete overview have a look at the [cheat sheet](graphics/lubridate.pdf).

A small fictitious dataset will be used here to explore some of the concepts. It can be found in file `data/time_series_drug_test.txt`.

```{r lubridate-data}
ts <- read.table("data/time_series_drug_test.txt", 
                 header = T,
                 sep = ";",
                 as.is = 2:3)
ts <- as_tibble(ts)
ts
```

If you simply want to show the timeseries as a plot, the date and time columns need to be combined first. The tidyr package has `unite()`, or do it with `mutate()`

```{r lubridate-unite}
ts %>% 
    unite("date_time", date, time, sep = "T") %>% 
    head(3)
```

As you can see, the original data columns are lost in this operation. Using `mutate()` and `paste()` does keep them intact however. I will take this approach because I want the old columns for demonstration purposes.

```{r lubridate-mutate}
ts <- ts %>% 
    mutate(date_time = paste(date, time, sep=" "))
head(ts, 3)
```

Let's reshuffle. I like it when my dependent variable comes last. 

```{r lubridate-select}
ts <- ts %>% select(subject, date, time, date_time, control, response)
head(ts, 3)
```

If I attempt to plot the response over time for each of the subjects I get into trouble:

```{r lubridate-ggplot, error=TRUE}
ggplot(data = ts, 
       mapping = aes(x = date, y = response, linetype = subject)) + 
    geom_line()
```

Also, when I would like to calculate something like the change per hour in response level over the different measurements this would be difficult indeed. Take a minute to think about this. Do you know any technique in base R that you've encountered so far that supports this?  
Actually there are some functions in baser R that can do this, but we'll stick to the tidyverse.

So this is the area of the `lubridate` package. I included a `library(lubridate)` statement at the top of this document so that's not needed anymore.

Let's start with reading in date and time variables. There are many dedicated (wrapper) functions for reading date and time data. The `date` variable has the format "**D**ay/**M**onth/**Y**ear" so the function `dmy()` should be fine here:

```{r lubridate-mutate-2}
ts <- ts %>% 
    mutate(date = dmy(ts$date))
ts %>% head(3)
```

As you can see, the type has changed into `date`. Let's try the plot again:

```{r lubridate-mutate-3, error=TRUE}
ggplot(data = ts, 
       mapping = aes(x = date, y = response, linetype = subject)) + 
    geom_line()
```


Now we're getting somewhere. The times are not taken into account yet, so it is simply mapped to 00:00 hours on that day. Also, we're not really interested in the specific date, we want to know the change from T = 0. We'll get to that.

Parsing time is the same as parsing the date, we have an "**H**our:**M**inute" format so the function to take is `hm()` (there is also an `hms()`):

```{r  lubridate-hm}
hm(ts$time)[1:5]
```

And yes, there is also an `dmy_hm()` function:

```{r lubridate-dmy-hs}
dmy_hm(ts$date_time)[1:5]
```
As long as the format is not too esoteric, lubridate will figure out the separator quite well.
 
If you want full control -or simply want to remember only one or two functions- you could take either one of `fast_strptime()` or `parse_date_time()`. Both take a **_format string_** as argument: a string in which you specify the way your date-time character data are structured The most used symbols are these (although you should really also look at the help page):

- **`b`** or **`B`** Abbreviated or full month name in the current locale. The C parser currently understands only English month names.
- **`d`** Day of the month as decimal number (01--31 or 0--31)
- **`H`** Hours as decimal number (00--24 or 0--24).
- **`I`** Hours as decimal number (01--12 or 1--12).
- **`m`** Month as decimal number (01--12 or 1--12). 
- **`M`** Minute as decimal number (00--59 or 0--59).
- **`p`** AM/PM indicator in the locale. Normally used in conjunction with I and not with H. 
- **`S`** Second as decimal number (00--61 or 0--61).
- **`y`** Year without century (00--99 or 0--99). In parse_date_time() also matches year with century (Y format).
- **`Y`** Year with century.

Function `parse_date_time()` is the most lenient of the two with respect to the **_format string_**:
 
```{r lubridate-parse-1}
parse_date_time(ts$date_time, "d.m-y H:M")[1:5] ##doesn't care I have a typo in my format string
ts <- ts %>% mutate(date_time = parse_date_time(ts$date_time, "d/m/y H:M"))
head(ts, 3)
```

The `fast_strptime()` is much more picky, but much faster. You specify the date(-time) format using`%` sign together with one of the symbols listed above:

```{r lubridate-parse-2}
fast_strptime('29/12/2019 07:44', '%d/%m/%Y %H:%M')
```

```{r lubridate-parse-3, eval = FALSE}
## When in tibble context, you need to set `lt = F`...
ts %>% 
    mutate(date_time = fast_strptime(date_time, '%d/%m/%Y %H:%M', lt = F))
```

I think sticking to the first will suffice.

### Current time

The functions `today()` and `now()` are your friends.


```{r lubridate-current}
##output will vary!
paste("the day is ", today())
paste("and to be really exact: ", now())
```


### Getting elements from date-time objects

Once you have the date-time object processed, you can look at the individual elements.

There are -again- many functions related to this. Have a look at the [cheat sheet](graphics/lubridate.pdf) for details. Here are the most-used components:

- **`date(x)`** Date component leaving the time out.
- **`year(x)`** The year. 
- **`month(x, label, abbr)`** The month. When called without other argument, you get the month number (e.g. 2). With the `label = TRUE` argument you will get text (e.g. Feb) and `abbr = FALSE` you get the full name of the month (February).
- **`day(x)`** The day of the month. 
- **`wday(x,label,abbr)`** The day of week. Label and abbr behaves the same as with `month()`.
- **`hour(x)`** The hour
- **`minute(x)`** Minutes.
- **`second(x)`** Seconds.
- **`week(x)`** Week of the year.

A use case demonstration: On what days were the samples taken of the times series used so far?

```{r lubridate-wday, results='hold'}

ts %>% 
    group_by(day_of_week = wday(date_time, label=T)) %>% 
    summarize(sample_count = n())
# or, shorter
#ts %>% group_by(day_of_week = wday(date_time, label=T)) %>% tally()
```

What was the mean response on the different times of the day:

```{r lubridate-hour}
ts %>% 
    group_by(hour_of_day = hour(date_time)) %>% 
    summarize(mean = mean(response))
```

And when split over the control/treatment group as well:

```{r lubridate-hour-2}
ts %>% 
    group_by(control, hour_of_day = hour(date_time)) %>% 
    summarize(mean = median(response),
              n = n())
```

Finally, let's close off with a plot

```{r lubridate-hour-3}
ggplot(data = ts, 
       mapping = aes(x = date_time, y = response, color = subject)) + 
    geom_line()
```

As practice, you could try to generate a response plot like this where subjects 1 to 3 have corrected values based on the non-treated group.

### Calculations with date-time

The example below shows the most basic calculation: the difference between two date-time objects.

```{r lubridate-calc}
ts %>% 
    group_by(subject) %>%
    select(subject, date_time) %>% 
    mutate(lagged = lag(date_time), 
           diff = date_time - lagged) 
```

You can define periods, durations and work with time intervals. However, this is outside the scope of this course.
For this more advanced time math you should refer to the cheat sheet or the official docs.

