# Text processing with regular expressions

## Introduction 

:::definitions
Pattern matching is the process of finding, locating, extracting and replacing patterns in character data that usually cannot be literally described.
:::

For instance, it is easy enough to look for the word "Chimpanzee" in a vector containing animal species names:

```{r}
animals = c("Chimpanzee", "Cow", "Camel")
animals == "Chimpanzee"
```

but what are you going to do if there are multiple variants of the word you are looking for? This?

```{r}
animals = c("Chimpanzee", "Chimp", "chimpanzee", "Camel")
animals == "Chimpanzee" | animals == "Chimp" | animals == "chimpanzee"
```

The solution here is not using literals, but to describe **_patterns_**.

Look at the above example. How would you describe a pattern that would correctly identify all Chimpanzee occurrences?

Is you pattern something like this?

_A letter C in upper-or lower case followed by 'himp' followed by nothing or 'anzee'_

In programming we use **_regular expressions_** or **_RegEx_** to describe such a pattern in a formal concise way:

`[Cc]himp(anzee)?`

And to apply such a pattern in R, we use one of several functions dedicated for this task. Here is one, `grepl()`, which returns `TRUE` if the regex matched the vector element.

```{r}
grepl("[Cc]himp(anzee)?", animals)
```

### Functions using regex {-}

There are  several functions dedicated to finding patters in character data. They differ in intent and output. Here is a listing.

- **finding** Does an element contain a pattern (TRUE/FALSE)? `grepl(pattern, string)`
- **locating** Which elements contain a pattern (INDEX)? `grep(pattern, string)`
- **extracting** Get the content of matching elements `grep(pattern, string, value = TRUE)`
- **replace** Replace the first occurrence of the pattern `sub(pattern, replacement, string)`
- **replace all** Replace all occurrences of the pattern `gsub(pattern, replacement, string)`

Note that the `stringr` package from the tidyverse has many user-friendly functions in this field as well. Two of them will be dealt with in the exercises.

### Regex syntax

A regular expression can be build out of any combination of

- **character sequences** - Literal character sequences such as 'chimp'
- **character classes** - A listing of possibilities for a single position.
    - Between brackets: `[adgk]` means 'a' or 'd' or 'g' or 'k'.
    - Use a hyphen to create a series: `[3-9]` means digits 3 through 9 and `[a-zA-Z]` means all alphabet characters.
    - Negate using `^`. `[^adgk]` means anything _but_ a, d, g or k.
    - A special case is the dot `.`: any character matches.
    - Many special character classes exist (digits, whitespaces etc). They are discussed in a later paragraph.
- **alternatives** - Are defined by the pipe symbol `|`: "OR"
- **quantifiers** - How many times the preceding block should occur. See next paragraph.
- **anchors** - `^` means matching at the start of a string. `$` means at the end.

An excellent cheat sheet from the RStudio website is also included
<a href="figures/RegExCheatsheet.pdf" target="_blank">here</a>

### Quantifiers

Use quantifiers to specify how many times a character or series of characters should occur.

- **`{n}`**: exactly `n` times
- **`{n, }`**: at least `n` times
- **`{ ,n}`**: at most `n` times
- **`{n, m}`**: at least `n` and at most `m` times.
- **`*`**: 0 or more times; same as `{0, }`
- **`+`**: 1 or more times; same as `{1, }`
- **`?`**: 0 or 1 time; same as `{0, 1}`

## Some examples

### Restriction enzymes {-}

This is the recognition sequence for the _HincII_ restriction endonuclease:

```
5'-GTYRAC-3'
3'-CARYTG-5'
```

Before reading on: how would you define a regular expression that is precisely describes this recognition sequence?

Molecular biology sequence ambiguity codes can be found
<a href="data/ambiguity-codes.html" target="_blank">here</a>

```{r}
HincII_rs <- "GT[CT][AG]AC"
sequences <- c("GTCAAC",
               "GTCGAC",
               "GTTGAC",
               "aGTTAACa",
               "GTGCAC")
grep(pattern = HincII_rs, x = sequences, value = TRUE)
```

### Dutch dates {-}

Here are some Dutch dates, in different accepted formats. The last two are not a correct notation.
Create a RegEx that will determine whether an element contains a Dutch date string.

```{r}
dates <- c("15/2/2019", "15-2-2019", "15-02-2019", "015/2/20191", "15/2/20191")
dateRegex <- "[0-9]{2}[/-][0-9]{1,2}[/-][0-9]{4}"
grep(pattern = dateRegex, x = dates, value = TRUE)
```

Why were the last two matched?
Because the pattern _is there_, albeit embedded in a longer string.
We have to **_anchor_** the pattern to be more specific.

## Anchoring

Using anchoring, you can make sure the string is not longer than you explicitly state:

```{r}
dates <- c("15/2/2019", "15-2-2019", "15-02-2019", "015/2/20191", "15/2/20191")
dateRegex <- "^[0-9]{2}[/-][0-9]{1,2}[/-][0-9]{4}$"
grep(pattern = dateRegex, x = dates, value = TRUE)
```

Now the date matching is correct.

## Metacharacters: Special character classes

Since patterns such as `[0-9]` occur so frequently, they have dedicated character classes such as `[[:digit:]]`. The most important other ones are

- **digits** `[[:digit:]]` or `\\d`: equivalent to `[0-9]`
- **alphabet characters** `[[:alpha:]]`: equivalent to `[a-zA-Z]`
- **lowercase characters** `[[:lower:]]`: equivalent to `[a-z]`
- **uppercase characters** `[[:upper:]]`: equivalent to `[A-Z]`
- **whitespace characters** `[[:space:]]` or `\\s`: Space, tab, vertical tab, newline, form feed, carriage return
- **punctuation characters** `[[:punct:]]`: One of !"#$%&’()*+,-./:;<=>?@[]^_`{|}~

(have a look at the cheat sheet for all)

Here is the same example, this time using these predefined character classes

```{r}
dates <- c("15/2/2019", "15-2-2019", "15-02-2019", "15022019", "15/2/20191")
dateRegex <- "[[:digit:]]{2}[/-]\\d{1,2}[/-]\\d{4}"
grep(pattern = dateRegex, x = dates, value = TRUE)
```

### Postal codes {-}

Here are some Dutch zip (postal) codes, in different accepted formats. The last two are not a correct notation.
Can you create a RegEx that will determine whether an element contains a Dutch zip code?

```{r}
zips <- c("1234 AA", "2345-BB", "3456CC", "4567 dd", "56789aa", "6789a_")
zips
```


### Prosite patterns {-}

<a href="https://prosite.expasy.org/" target="_blank">Prosite</a> is a database of amino acid sequence motifs. One of them is the Histidine Triad profile (PDOC00694).

```
[NQAR]-x(4)-[GSAVY]-x-[QFLPA]-x-[LIVMY]-x-[HWYRQ]-
[LIVMFYST]-H-[LIVMFT]-H-[LIVMF]-[LIVMFPT]-[PSGAWN]
```

- Write this down as a RegEx
- Was that efficient? Using the `gsub()` function, can you convert it in a RegEx using code? It may take several iterations. Was that efficient?
- Next, use an appropriate function to find if, and where, this pattern is located within the sequences in file `data/hit_proteins.txt` (<a href="data/hit_proteins.txt" target="_blank">here</a>)

Amino Acid codes and Prosite pattern encoding can be found
<a href="data/ambiguity-codes.html" target="_blank">here</a>


## Locating and Extracting

This is outside the scope of this first acquaintance. The `stringr` package from the tidyverse is much better at this than the base R functions. One of the exercises introduces this topic for the eager ones among you.

